Search for Spatially Extended \fermi-LAT Sources Using Two Years of Data

### 2011-11-02 extended_catalog/paper/extended_catalog.tex
updating text

% New text (from v3.0 to v?.?):
%  * Better clarification about previous gtlike validation of extended source analysis.
%  * Improve MC null distribution tests by using 8 bins per decade + perofrm new test with E>10GeV.

@@

\title{Search for Spatially Extended \fermi-LAT Sources Using Two Years of Flight
Data}
\shorttitle{Search for Extended LAT Sources}

@@

\author{
J.~Lande\altaffilmark{2,1}, 
S.~Funk\altaffilmark{2,1}, 
M.~Ackermann\altaffilmark{3,1}, 
%A.~Allafort\altaffilmark{2},
%M.-H.~Grondin\altaffilmark{53},
% Keith Bechtol
% Marshall Roth?
% Marianne Lemoine-Goumard
% Francesco Giordano
% Toby Burnett
% Jean Ballet
% Matthew Kerr
\ldots
\altaffiltext{1}{Corresponding authors: M.~Ackermann, markus.ackermann@desy.de; S.~Funk, funk@slac.stanford.edu; J.~Lande, joshualande@gmail.com.}
\altaffiltext{2}{W. W. Hansen Experimental Physics Laboratory, Kavli Institute for Particle Astrophysics and Cosmology, Department of Physics and SLAC National Accelerator Laboratory, Stanford University, Stanford, CA 94305, USA}
\altaffiltext{3}{Deutsches Elektronen Synchrotron DESY, D-15738 Zeuthen, Germany}
}

@@

A number of astrophysical source classes including supernova remnants
(SNRs), pulsar wind nebulae (PWNe), molecular clouds, normal galaxies,
and galaxy clusters are expected to be spatially resolvable at
\gev energies.  Dark matter satellites are also hypothesized to
be spatially extended.  The Large Area Telescope
(LAT) on the {\em \fermi Gamma-ray Space Telescope} (\fermi) has
detected seven SNRs which are significantly extended
at \gev energies: 
W51C, W30, 
IC443, W28, W44, RX\,J1713.7$-$3946,
and the Cygnus Loop
\citep{w51c,w30_lat,ic443,w28,w44,rx_j1713_lat,cygnus_loop_lat}. In addition, three 
PWNe were detected as being extended: MSH\,15$-$52, Vela~X, and
HESS\,J1825$-$137 \citep{msh1552,velax,fermi_hess_j1825}. Two
close-by galaxies, the Large and Small Magellanic Clouds, and one
radio galaxy, Centarus A, were spatially resolved at \gev energies
\citep{lmc,smc,cen_a_lat}.  A number of additional sources detected
at \gev energies are positionally coincident with sources that exhibit
extension at other wavelengths, large enough to be spatially
resolvable by the LAT at \gev energies.

The current generation of air Cherenkov detectors have
made it apparent that many sources can be spatially resolved
at even higher energies.
Most
prominent was a survey of the Galactic plane using the High Energy
Stereoscopic System (H.E.S.S) which reported 14 spatially extended
sources with extensions varying from $\sim0\fdg1$ to $\sim0\fdg25$
\citep{hess_plane_survey}.  In fact, within our Galaxy only very few
sources detected at \tev energies (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61$-$303 \citep{MAGICLSI, VERITASLSI},
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes})
have no detectable extension.  High-energy $\gamma$-rays from
these sources are produced by the decay of $\pi^0$s produced by hadronic
interactions with interstellar matter and by relativistic electrons
due to Inverse Compton (IC) scattering and Bremsstrahlung radiation.  
It is plausible that the \gev and
\tev emission from these sources originates from the same population of
high-energy particles and so at least some of these sources should be
detected at \gev energies.  Studying 
these \tev sources at \gev energies would help to
determine the emission mechanisms producing these high energy photons.

The LAT is a pair conversion telescope that has been surveying the
$\gamma$-ray sky since 2008 August.  The LAT has broad energy coverage
(20 \mev to $>300$ \gev), wide field of view ($\sim 2.4$ sr), and large
effective area ($\sim 8000\ \cm^2$ at $>1 \gev$) 
Additional information about the performance of the LAT can be found in
\cite{atwood_LAT_mission}.
Using 2 years of all-sky surveying data, the LAT Collaboration published
a catalog of 1873
sources significantly detected at \gev energies called 2FGL \citep{second_cat}.
The possible counterparts of many of these sources can be spatially resolved
when observed at other frequencies but detecting the spatial extension
of these sources at \gev energies is difficult because the size of the
point-spread function (PSF) of the LAT is comparable to the typical size
of many of these sources.

@@

\subsection{The \pointlike Package}

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT.
The Poisson likelihood to find
the observing counts is maximized given a parametrized spatial and
spectral model of the source and its surrounding region.  The sky is
divided into cubes of space and energy using the healpix representation
of the sky \citep{healpix} and the likelihood is maximized over all bins
in a region.  The extension of a source can be modeled by a geometric
shape (e.g. a disk or a two-dimensional Gaussian) and the source's position, extension,
and spectrum can be simultaneously fit.

@@

In \pointlike, it is assumed that the spatial and spectral model
of an extended source are separable, i.e. that the source model
$M(l,b,E)=S(l,b)\times X(E)$ where $S(l,b)$ is the spatial distribution
and $X(E)$ is the spectral distribution.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a function
of energy) and uses the \minuit library \citep{minuit_documentation}
to maximize the likelihood by simultaneously varying the position,
extension, and spectrum of the source.  As will be described in
Section~\ref{monte_carlo_validation}, simultaneously fitting the
position, extension, and spectrum is important to maximize
the statistical significance of the detection of extension of a source.
To avoid projection effects, the source's longitude and latitude are
not directly fit but instead the source's displacement in a rotated
reference frame.

@@

Figure~\ref{all_sensitivity} shows the flux threshold as a function of
source extension for different background levels ($1\times$, $10\times$,
and $100\times$ the nominal background), different spectral indices,
and two different energy ranges (1 \gev to 100 \gev and 10 \gev to
100 \gev).  The detection threshold is higher for sources in regions of
higher background.  When studying sources only at energies above 1 \gev,
the LAT detection threshold (defined as the 1 \gev to 100 \gev flux at
which $\langle\tsext\rangle=16$) depends less strongly on the source's
spectral index. 
The index dependence of the detection threshold is
even weaker when considering only photons with energies above 10 \gev
because the PSF changes little from 10 \gev to 100 \gev.
Overlaid on Figure~\ref{all_sensitivity} are the LAT detected extended
sources that will be discussed in Sections~\ref{validate_known} and
\ref{new_ext_srcs_section}.  The extension thresholds are tabulated in
Table~\ref{all_sensitivity_table}.

@@

We did this study first in the energy range from 1 \gev to 100 \gev by
simulating extended sources of flux $4\times10^{-9} \phflux$ integrated
from 1 \gev to 100 \gev with a power-law spectral model with
spectral index 2.  This flux was picked to be representative of the
new extended sourced that were discovered in the following analysis
when looking in the 1 \gev to 100 \gev energy range
(see Section~\ref{new_ext_srcs_section}).
We simulated these sources using a uniform surface brightness profile
with extensions varying
up to $1\degree$.  Figure~\ref{confusion_plot}a shows 
the distribution of \tsinc and \tsext as a
function of the simulated extension of the source
for 200 statistically independent simulations.

Figure~\ref{confusion_plot}c shows the same analysis but when
fitting simulations of two point-like sources each with half of the
flux of the spatially extended source and with the same spectral index
as the extended source.
Finally, figure~\ref{confusion_plot}e shows the same analysis with each
point-like source having the same flux but different spectral indices.
One point-like source had a spectral index of 1.5 and the other an index
of 2.5.

@@

The distributions also shows that for simulated
extended sources with typical
sizes ($\sigma\sim0\fdg5$), one can often obtain almost as good an
increase in likelihood
when fitting the extended source as two point-like sources.
This is because although
two point-like sources represent an incorrect spatial model, 
the two point-like sources have four additional degrees
of freedom (two spatial and two spectral parameters)
than one extended source (one extension parameter)
and can therefore easy model
the extended source and statistical fluctuations in the data.
This effect is most pronounced when using photons with energies between 1
\gev and 100 \gev where the PSF is even larger.  
On the other hand,
when testing the two simulated point-like sources for extension our test
did a better job at separating the distribution of \tsinc and \tsext
and found that \tsinc was typically much larger than \tsext.
This is because the one additional extension parameter is less able to
model the second source or statistical fluctuations.

@@

Finally, figure~\ref{index_difference} shows the absolute value of the
difference of the fit spectral indices for the case of fitting simulated
spatially extended sources as two point-like sources. The plot compares
the fit in the 1 \gev to 100 \gev energy range and in the 10 \gev to
100 \gev energy range. An error on the difference in spectral index is
calculated by adding in quadrature the statistical errors on the spectral
indices.  This plot shows that there can be a large difference in the
spectral index of the two fit point-like sources which is often greater
than the statistical errors on the difference in spectral index. This is
because the fit values the errors on an incorrect model are not expected
to be meaningful. Therefore, the difference in spectral index when fitting
an extended source candidate as two point-like sources can not be easily
used to help determine whether a source is actually spatially extended
instead of being the emission of two nearby point-like sources.

@@

Unless explicitly
mentioned, we used the same background model as 2FGL to represent the
Galactic diffuse, isotropic, and Earth limb emission.  To account for
possible residuals in the diffuse emission model, the galactic emission
was scaled by a power-law and the isotropic component's normalization
was left free.  
Unless explicitly mentioned,
we used all 2FGL sources within $15\degree$ of our source as our list
of background sources and we refit the spectral parameters of all sources
within $2\degree$ of the source.

@@

Of the 885 clean AGN, we selected the 783 of these 2FGL sources which
were significantly-detected above 1 \gev and fit each of them for extension.
The cumulative density of \tsext for these AGN compared to the
$\chi^2_1/2$ distribution of Equation~\ref{ts_ext_distribution}
is
shown in Figure~\ref{agn_ts_ext}. 
The \tsext distribution for AGN shows good agreement with the
theoretical distribution, but two sources had $\tsext>10$.  One was due
to the incorrect removal of a nearby catalog sources from
the sky model (see
Section~\ref{extended_source_search_method}) and the other was due to a
failure of convergence of the point-like source hypothesis.  This result demonstrates
that we can use \tsext as a measure of the statistical significance of
the detection of the extension of a source.

@@

As further validation of our extended source method for studying
spatially extended sources, we reanalyzed the twelve spatially extended
sources which were included in 2FGL \citep{second_cat}.  Even though
each of these sources required a dedicated analysis and they
were fit with a variety of spatial models, it is valuable to show that
these sources are significantly extended using our systematic extension
method that assumed a radially-symmetric uniform surface
brightness profile.  On the other hand, for some of
these sources a uniform disk spatial model does not well describe the
observed extended emission and so the dedicated LAT publications provides
a qualitatively better model these sources.

@@

% 3 PWN
Three PWNe, MSH\,15$-$52, Vela X, and HESS\,J1825$-$137 were
identified in 2FGL \citep{msh1552,velax,fermi_hess_j1825}.  
HESS\,J1825$-$137 was significantly detected using photons
with energies between 10 \gev and 100 \gev.
To improve our analysis of this source, we had to remove
from our background model 
2FGL\,J1823.1$-$1338c which was included in 2FGL to
represent part of the extended emission.
To avoid confusion with the nearby bright pulsar PSR\,J1509$-$5850, MSH\,15$-$52 had
to be
analyzed at high energies.  Using photons with energies above 10 \gev,
we fit the extension of MSH\,15$-$52 to be consistent with the published
size but with an extension significance of \tsext=6.5.

% failed to detect vela x + cen a
Our analysis was unable to resolve Vela X which would have required first
removing the pulsed photons from the Vela pulsar which was beyond the
scope of this paper.  Our analysis also failed to detect a significant
extension for the Centaurus A Lobes because
the source's emission is significantly different from a uniform
radially-symmetric surface brightness profile \citep{cen_a_lat}.

@@

We estimated a second systematic error on the extension of a source
due to uncertainty in our model of the Galactic diffuse emission by
using an alternative model which takes as input templates
calculated by
GALPROP\footnote{GALPROP is a software package for calculating the
Galactic $\gamma$-ray emission based on a model of cosmic-ray propagation
in the Galaxy \citep{galprop1998,galprop2011}. 
See also \url{http://galprop.stanford.edu/} for details.} 
but then fits each template locally in
the surrounding region.
The particular GALPROP model that was used as input is described in
the LAT analysis of the isotropic diffuse
emission \citep{isotropic_lat}.  
The intensities of various components
of the Galactic diffuse emission were fitted individually using a
spatial distribution of the intensities within the ROI as predicted by
the model.  We distinguished contributions from CR interactions with the
molecular hydrogen, the atomic+ionized hydrogen, residual gas traced
by dust \citep{isabelle_dark_gass}, and the interstellar radiation
field. We further split the contributions from interactions with molecular
and atomic hydrogen to the Galactic diffuse emission according to the
distance from the Galactic center in which they are produced. Hence, we
replaced the standard diffuse emission model by 18 individually fitted
templates to describe individual components of the diffuse emission.
A similar crosscheck was used in the LAT Collaboration's analysis of 
RX\,J1713.7$-$3946 \citep{rx_j1713_lat}.

@@

But this is infeasible in the galactic plane where we most likely expect
to discover new spatially extended sources.  In the galactic plane,
this analysis is challenged by our imperfect model of the diffuse
emission and by issues related to source confusion.  The Monte Carlo
study in Section~\ref{dual_localization_method}
showed that the overall likelihood would greatly increase when fitting
a spatially extended source as two point-like sources so we expect
that spatially extended LAT sources are modeled in 2FGL as
multiple point-like sources. Furthermore, the position of other nearby sources
in the region near an extended source could be biased by not correctly
modeling the
sources' extension.  Furthermore, 2FGL produced a
list of sources significant at energies above 100 \mev whereas we are
most sensitive to spatial extension at higher energies.  Therefore,
we expect that at higher energies our analysis would be complicated
by 2FGL sources no longer significant and by 2FGL
sources whose position was biased by diffuse emission at lower energies.

To account for these issues, we first produces a large list of
possibly extended sources employing a very liberal search criteria
and then refined the analysis of the promising candidates on a case
by case basis.  Our strategy tested all point-like 2FGL sources for
extension assuming they had a uniform radially-symmetric uniform surface
brightness profile and a power-law spectral model.  Although not all
extended sources are expected to have a shape very similar to a uniform
surface brightness profile, Section~\ref{compare_source_size} showed that
for many spatially extended sources the wide PSF of the LAT and limited
statistics makes this a reasonable approximation.  Additionally, our Monte
Carlo study and AGN validation (Sections~\ref{monte_carlo_validation}
and~\ref{test_2lac_sources}) demonstrate that we can use this shape
for a statistically rigorous test. Because this spatial model only adds
one additional degree of freedom, this test should be very sensitive.
On the other hand, choosing this spatial model could bias us against
finding extended sources that are not well described by uniform emission
such as shell-type supernova remnants.

@@

When we applied this test to the 1861 point-like sources in 2FGL, our
search found 117 extended source candidates in the 1 \gev to 100 \gev
energy range and 11 extended source candidates in the 10 \gev to 100
\gev energy range. Most of the extended sources flagged above 10 \gev were
also flagged above 1 \gev.  Many of these situations represent multiple
point-like 2FGL sources all fitting the same emission region. For example,
the sources 2FGL\,J1630.2-4752, 2FGL\,J1632.4-4753c 2FGL\,J1634.4-4743c,
and 2FGL\,J1636.3-4740c were all found to be spatially extended in the
10 \gev to 100 \gev energy range even though they all fit to similar
positions and sizes.  For these situations, we manually discarded all
but one of the 2FGL sources.

Similarly, many of these sources were confused with nearby
point-like sources or influence by large-scale residuals in the
diffuse emission.  To help determine which of these fits represented
truly extended sourced and in which of these fits the extension was
influenced by source confusion and diffuse emission, we generated a
series of diagnostic plots.  For each candidate, we generated a map
of the residual test statistic by adding a new source of spectral index 2 into the
region at each position and finding the increase in likelihood when fitting
its flux. Figure~\ref{res_tsmaps} shows this map around the
most significantly extended source IC443 when it is modeled both as a
point-like sources and an extended source.  The residual test statistic
map indicates
that the spatially extended model for IC443 is a significantly better
description of the observed photons and that there is not significant
residual in region after modeling the source as being spatially extended.
We also generated maps of the sum of all counts within a given distance of
the source and compared it to the model predictions assuming the emission
originated from a point-like source.  An example radial integral plot
is shown for the extended source IC443 in Figure~\ref{four_plots_ic443}.
For each source, we also made diffuse-emission-subtracted smoothed counts
maps (show for IC443 in Figure~\ref{four_plots_ic443}).

>>>>

% New text (from v3.0 to v4.0):
%  * Improve MC null distribution tests by using 8 bins per decade.
%  * New study of null hypothesis distribution fitting only above E>10GeV.
%  * Regenerate source confusion plot to show TS_ext-TS2pts
%  * Remove figure 8 from draft v3.0  (the one describing the spectral index difference of the two point 
%    source fit of extended sources) since everybody asked me to.
%  * Rerun of agn analysis to fix sources that failed to converge + using 8bins per decade + using 2FGL skymodel.
%  * For new source anlaysis:
%  ** use 8 insetad of 4 bins per decade
%  ** Freeze instead of delete insignificant 2FGL sources. This should make the text clearer because sources are only
%     deleted if they are part of the extended emission.
%  ** Add two nearby point-like sources to analysis of HESS J1632-478.
%  ** Redo analysis of RX J1713 to be E>10GeV to avoid issues of diffuse uncertainty at low energy.
%  * Better clarification about previous gtlike validation of extended source analysis + Citation of Francesco's work.

@@

\title{Search for Spatially Extended \fermi-LAT Sources Using Two Years of Data}
\shorttitle{Search for Extended LAT Sources}

@@

\author{
J.~Lande\altaffilmark{1,2}, 
M.~Ackermann\altaffilmark{3,4}, 
A.~Allafort\altaffilmark{1}, 
K.~Bechtol\altaffilmark{1}, 
T.~H.~Burnett\altaffilmark{5}, 
A.~Drlica-Wagner\altaffilmark{1}, 
S.~Funk\altaffilmark{1,6}, 
F.~Giordano\altaffilmark{7,8}, 
M.-H.~Grondin\altaffilmark{9,10}, 
M.~Kerr\altaffilmark{1}, 
M.~Lemoine-Goumard\altaffilmark{11,12}
}
\altaffiltext{1}{W. W. Hansen Experimental Physics Laboratory, Kavli Institute for Particle Astrophysics and Cosmology, Department of Physics and SLAC National Accelerator Laboratory, Stanford University, Stanford, CA 94305, USA}
\altaffiltext{2}{email: joshualande@gmail.com}
\altaffiltext{3}{Deutsches Elektronen Synchrotron DESY, D-15738 Zeuthen, Germany}
\altaffiltext{4}{email: markus.ackermann@desy.de}
\altaffiltext{5}{Department of Physics, University of Washington, Seattle, WA 98195-1560, USA}
\altaffiltext{6}{email: funk@slac.stanford.edu}
\altaffiltext{7}{Dipartimento di Fisica ``M. Merlin" dell'Universit\`a e del Politecnico di Bari, I-70126 Bari, Italy}
\altaffiltext{8}{Istituto Nazionale di Fisica Nucleare, Sezione di Bari, 70126 Bari, Italy}
\altaffiltext{9}{Max-Planck-Institut f\"ur Kernphysik, D-69029 Heidelberg, Germany}
\altaffiltext{10}{Landessternwarte, Universit\"at Heidelberg, K\"onigstuhl, D 69117 Heidelberg, Germany}
\altaffiltext{11}{Universit\'e Bordeaux 1, CNRS/IN2p3, Centre d'\'Etudes Nucl\'eaires de Bordeaux Gradignan, 33175 Gradignan, France}
\altaffiltext{12}{Funded by contract ERC-StG-259391 from the European Community}

@@

A number of astrophysical source classes including supernova remnants
(SNRs), pulsar wind nebulae (PWNe), molecular clouds, normal galaxies,
and galaxy clusters are expected to be spatially resolvable at
\gev energies.  Dark matter satellites are also hypothesized to
be spatially extended.  The Large Area Telescope
(LAT) on the {\em \fermi Gamma-ray Space Telescope} (\fermi) has
detected seven SNRs which are significantly extended
at \gev energies: 
W51C, W30, 
IC443, W28, W44, RX\,J1713.7$-$3946,
and the Cygnus Loop
\citep{w51c,w30_lat,ic443,w28,w44,rx_j1713_lat,cygnus_loop_lat}. In addition, 
three extended PWN have been detected: MSH\,15$-$52, Vela~X, and
HESS\,J1825$-$137 \citep{msh1552,velax,fermi_hess_j1825}. Two
nearby galaxies, the Large and Small Magellanic Clouds, and one
radio galaxy, Centaurus A, were spatially resolved at \gev energies
\citep{lmc,smc,cen_a_lat}.  A number of additional sources detected
at \gev energies are positionally coincident with sources that exhibit
extension at other wavelengths, large enough to be spatially
resolvable by the LAT at \gev energies.

The current generation of air Cherenkov detectors have
made it apparent that many sources can be spatially resolved
at even higher energies.
Most
prominent was a survey of the Galactic plane using the High Energy
Stereoscopic System (H.E.S.S) which reported 14 spatially extended
sources with extensions varying from $\sim0\fdg1$ to $\sim0\fdg25$
\citep{hess_plane_survey}.  In fact, within our Galaxy only very few
sources detected at \tev energies, most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61$-$303 \citep{MAGICLSI, VERITASLSI},
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes},
have no detectable extension.  High-energy $\gamma$-rays from
these sources are produced by the decay of $\pi^0$s produced by hadronic
interactions with interstellar matter and by relativistic electrons
due to Inverse Compton (IC) scattering and Bremsstrahlung radiation.  
It is plausible that the \gev and
\tev emission from these sources originates from the same population of
high-energy particles and so at least some of these sources should be
detected at \gev energies.  Studying 
these \tev sources at \gev energies would help to
determine the emission mechanisms producing these high energy photons.

The LAT is a pair conversion telescope that has been surveying the
$\gamma$-ray sky since 2008 August.  The LAT has broad energy coverage
(20 \mev to $>300$ \gev), wide field of view ($\sim 2.4$ sr), and large
effective area ($\sim 8000\ \cm^2$ at $>1$ \gev) 
Additional information about the performance of the LAT can be found in
\cite{atwood_LAT_mission}.
Using 2 years of all-sky surveying data, the LAT Collaboration published
a catalog of 1873
sources significantly detected at \gev energies called 2FGL \citep{second_cat}.
The possible counterparts of many of these sources can be spatially resolved
when observed at other frequencies but detecting the spatial extension
of these sources at \gev energies is difficult because the size of the
point-spread function (PSF) of the LAT is comparable to the typical size
of many of these sources.

@@

\subsection{Modeling Extended Sources in the \pointlike Package}

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT.
The Poisson likelihood to find
the observing counts is maximized given a parametrized spatial and
spectral model of the source and its surrounding region.  The sky is
divided into cubes of space and energy using the healpix representation
of the sky \citep{healpix} and the likelihood is maximized over all bins
in a region.  The extension of a source can be modeled by a geometric
shape (e.g. a disk or a two-dimensional Gaussian) and the position, extension,
and spectrum of the source
can be simultaneously fit.

@@

In \pointlike, it is assumed that the spatial and spectral model
of an extended source are separable, i.e. that the source model
$M(l,b,E)=S(l,b)\times X(E)$ where $S(l,b)$ is the spatial distribution
and $X(E)$ is the spectral distribution.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a function
of energy) and uses the \minuit library \citep{minuit_documentation}
to maximize the likelihood by simultaneously varying the position,
extension, and spectrum of the source.  As will be described in
Section~\ref{monte_carlo_validation}, simultaneously fitting the
position, extension, and spectrum is important to maximize
the statistical significance of the detection of extension of a source.
To avoid projection effects, the longitude and latitude 
of the source are
not directly fit but instead the displacement of the source in a rotated
reference frame.

@@

Figure~\ref{all_sensitivity} shows the flux threshold as a function of
source extension for different background levels ($1\times$, $10\times$,
and $100\times$ the nominal background), different spectral indices,
and two different energy ranges (1 \gev to 100 \gev and 10 \gev to
100 \gev).  The detection threshold is higher for sources in regions of
higher background.  When studying sources only at energies above 1 \gev,
the LAT detection threshold (defined as the 1 \gev to 100 \gev flux at
which $\langle\tsext\rangle=16$) depends less strongly on the 
spectral index of the source. 
The index dependence of the detection threshold is
even weaker when considering only photons with energies above 10 \gev
because the PSF changes little from 10 \gev to 100 \gev.
Overlaid on Figure~\ref{all_sensitivity} are the LAT detected extended
sources that will be discussed in Sections~\ref{validate_known} and
\ref{new_ext_srcs_section}.  The extension thresholds are tabulated in
Table~\ref{all_sensitivity_table}.

@@

We did this study first in the energy range from 1 \gev to 100 \gev by
simulating extended sources of flux $4\times10^{-9} \phflux$ integrated
from 1 \gev to 100 \gev with a power-law spectral model with
spectral index 2.  This flux was picked to be representative of the
new extended sourced that were discovered in the following analysis
when looking in the 1 \gev to 100 \gev energy range
(see Section~\ref{new_ext_srcs_section}).
We simulated these sources using a uniform surface brightness profile
with extensions varying
up to $1\degree$.  
\newtext{
Figure~\ref{confusion_plot}a shows 
the distribution of $\tsext-\tsinc$ as a
function of the simulated extension of the source
for 200 statistically independent simulations.
}

Figure~\ref{confusion_plot}c shows the same plot but when fitting
simulations of two point-like sources each with half of the flux of
the spatially extended source and with the same spectral index as the
extended source.  Finally, figure~\ref{confusion_plot}e shows the same
plot with each point-like source having the same flux but different
spectral indices.  One point-like source had a spectral index of 1.5
and the other an index of 2.5.

@@

The distributions also shows that for simulated
extended sources with typical
sizes ($\sigma\sim0\fdg5$), one can often obtain almost as good an
increase in likelihood
when fitting the extended source as two point-like sources ($\tsinc\sim\tsext$).
This is because although
two point-like sources represent an incorrect spatial model, 
the two point-like sources have four additional degrees
of freedom (two spatial and two spectral parameters)
than one extended source (one extension parameter)
and can therefore easy model
the extended source and statistical fluctuations in the data.
This effect is most pronounced when using photons with energies between 1
\gev and 100 \gev where the PSF is even larger.  
On the other hand,
when testing the two simulated point-like sources for extension our test
did a better job at separating the distribution of \tsinc and \tsext
and found that \tsinc was larger than \tsext.
This is because the one additional extension parameter is less able to
model the second source or statistical fluctuations.

@@

@@

Unless explicitly
mentioned, we used the same background model as 2FGL to represent the
Galactic diffuse, isotropic, and Earth limb emission.  To account for
possible residuals in the diffuse emission model, the galactic emission
was scaled by a power-law and the normalization
of the isotropic component
was left free.  
Unless explicitly mentioned,
we used all 2FGL sources within $15\degree$ of our source as our list
of background sources and we refit the spectral parameters of all sources
within $2\degree$ of the source.

@@

\newtext{
Of the 885 clean AGN, we selected the 733 of these 2FGL sources which
were significantly detected above 1 \gev and fit each of them for
extension.  The cumulative density of \tsext for these AGN is compared
to the $\chi^2_1/2$ distribution of Equation~\ref{ts_ext_distribution}
and is shown in Figure~\ref{agn_ts_ext}.  The \tsext distribution
for the AGN shows reasonable agreement with the theoretical
distribution and no AGN was found to be significantly extended
($\tsext>16$).  The observed discrepancy from the theoretical
distribution is likely due to small systematics
in our model of the LAT PSF and the Galactic diffuse emission (see
Section~\ref{systematic_errors_on_extension}).  The discrepancy could
also in a few cases be due to confusion with a nearby undetected source.
The overall agreement with the expected distribution demonstrates that
we can use \tsext as a measure of the statistical significance of the
detection of the extension of a source.
}

@@

As further validation of our extended source method for studying
spatially extended sources, we reanalyzed the twelve spatially extended
sources which were included in 2FGL \citep{second_cat}.  Even though
each of these sources required a dedicated analysis and they
were fit with a variety of spatial models, it is valuable to show that
these sources are significantly extended using our systematic 
method that assumed a radially-symmetric uniform surface
brightness profile.  On the other hand, for some of
these sources a uniform disk spatial model does not well describe the
observed extended emission and so the dedicated LAT publications provides
a qualitatively better model of these sources.

@@

% 3 PWN
Three PWNe, MSH\,15$-$52, Vela X, and HESS\,J1825$-$137 were
identified in 2FGL \citep{msh1552,velax,fermi_hess_j1825}.  
HESS\,J1825$-$137 was significantly detected using photons
with energies between 10 \gev and 100 \gev.
To improve our analysis of this source, we had to remove
from our background model 
2FGL\,J1823.1$-$1338c which was included in 2FGL to
represent part of the extended emission.
To avoid confusion with the nearby bright pulsar PSR\,J1509$-$5850, MSH\,15$-$52 had
to be
analyzed at high energies.  Using photons with energies above 10 \gev,
we fit the extension of MSH\,15$-$52 to be consistent with the published
size but with an extension significance of \tsext=6.6.

% failed to detect vela x + cen a
Our analysis was unable to resolve Vela X which would have required first
removing the pulsed photons from the Vela pulsar which was beyond the
scope of this paper.  Our analysis also failed to detect a significant
extension for the Centaurus A Lobes because
the shape of the source is significantly different from a uniform
radially-symmetric surface brightness profile \citep{cen_a_lat}.

@@

We estimated a second systematic error on the extension of a source
due to uncertainty in our model of the Galactic diffuse emission by
using an alternative model which takes as input templates
calculated by
GALPROP\footnote{GALPROP is a software package for calculating the
Galactic $\gamma$-ray emission based on a model of cosmic-ray propagation
in the Galaxy \citep{galprop1998,galprop2011}. 
See also \url{http://galprop.stanford.edu/} for details.} 
but then fits each template locally in
the surrounding region.
The particular GALPROP model that was used as input is described in
the LAT analysis of the isotropic diffuse
emission \citep{isotropic_lat}.  
The intensities of various components
of the Galactic diffuse emission were fitted individually using a
spatial distribution of the intensities within the ROI as predicted by
the model.  We distinguished contributions from CR interactions with the
molecular hydrogen, the atomic+ionized hydrogen, residual gas traced
by dust \citep{isabelle_dark_gass}, and the interstellar radiation
field. We further split the contributions from interactions with molecular
and atomic hydrogen to the Galactic diffuse emission according to the
distance from the Galactic center in which they are produced. Hence, we
replaced the standard diffuse emission model by 18 individually fitted
templates to describe individual components of the diffuse emission.
A similar crosscheck was used in an analysis of RX\,J1713.7$-$3946 
by the LAT Collaboration \citep{rx_j1713_lat}.

@@

But this is infeasible in the galactic plane where we are most likely expect
to discover new spatially extended sources.  In the galactic plane,
this analysis is challenged by our imperfect model of the diffuse
emission and by issues related to source confusion.  The Monte Carlo
study in Section~\ref{dual_localization_method}
showed that the overall likelihood would greatly increase when fitting
a spatially extended source as two point-like sources so we expect
that spatially extended LAT sources are modeled in 2FGL as
multiple point-like sources. Furthermore, the position of other nearby sources
in the region near an extended source could be biased by not correctly
modeling the
sources' extension.  Furthermore, 2FGL produced a
list of sources significant at energies above 100 \mev whereas we are
most sensitive to spatial extension at higher energies.  Therefore,
we expect that at higher energies our analysis would be complicated
by 2FGL sources no longer significant and by 2FGL
sources whose position was biased by diffuse emission at lower energies.

To account for these issues, we first produced a large list of
possibly extended sources employing a very liberal search criteria
and then refined the analysis of the promising candidates on a case
by case basis.  Our strategy tested all point-like 2FGL sources for
extension assuming they had a uniform radially-symmetric uniform surface
brightness profile and a power-law spectral model.  Although not all
extended sources are expected to have a shape very similar to a uniform
surface brightness profile, Section~\ref{compare_source_size} showed that
for many spatially extended sources the wide PSF of the LAT and limited
statistics makes this a reasonable approximation.  Additionally, our Monte
Carlo study and AGN validation (Sections~\ref{monte_carlo_validation}
and~\ref{test_2lac_sources}) demonstrate that we can use this shape
for a statistically rigorous test. Because this spatial model only adds
one additional degree of freedom, this test should be very sensitive.
On the other hand, choosing this spatial model biases us against
finding extended sources that are not well described by uniform emission
such as shell-type supernova remnants.

@@

When we applied this test to the 1861 point-like sources in 2FGL, our
search found 117 extended source candidates in the 1 \gev to 100 \gev
energy range and 11 extended source candidates in the 10 \gev to 100
\gev energy range. Most of the extended sources flagged above 10 \gev were
also flagged above 1 \gev and many of these situations represent multiple
point-like 2FGL sources all fitting the same emission region. For example,
the sources 2FGL\,J1630.2-4752, 2FGL\,J1632.4-4753c 2FGL\,J1634.4-4743c,
and 2FGL\,J1636.3-4740c were all found to be spatially extended in the
10 \gev to 100 \gev energy range even though they all fit to similar
positions and sizes.  For these situations, we manually discarded all
but one of the 2FGL sources.

Similarly, many of these sources were confused with nearby
point-like sources or influence by large-scale residuals in the
diffuse emission.  To help determine which of these fits found
truly extended sourced and when the extension was
influenced by source confusion and diffuse emission, we generated a
series of diagnostic plots.  For each candidate, we generated a map
of the residual test statistic by adding a new point-like source of spectral index 2 into the
region at each position and finding the increase in likelihood when fitting
its flux. Figure~\ref{res_tsmaps} shows this map around the
most significantly extended source IC443 when it is modeled both as a
point-like sources and an extended source.  The residual test statistic
map indicates
that the spatially extended model for IC443 is a significantly better
description of the observed photons and that there is not significant
residual in the region after modeling the source as being spatially extended.
We also generated maps of the sum of all counts within a given distance of
the source and compared it to the model predictions assuming the emission
originated from a point-like source.  An example radial integral plot
is shown for the extended source IC443 in Figure~\ref{four_plots_ic443}.
For each source, we also made diffuse-emission-subtracted smoothed counts
maps (shown for IC443 in Figure~\ref{four_plots_ic443}).

It was very clear by visual inspection that in many
cases our analysis was strongly influence by large-scale residual in the
diffuse emission and hence the extension measure was unreliable.  This was
especially true in our analysis of sources in the 1 \gev to 100 \gev
energy range.  An example of such a case is 2FGL\,J1856.2+0450c analyzed
in the 1 \gev to 100 \gev energy range. Figure~\ref{example_bad_fit}
shows a diffuse-emission-subtracted smoothed counts map for
this source with the best
fit extension of the source overlaid. There appears to be large-scale
residual in the diffuse emission in this region along the Galactic plane.
As a result, 2FGL\,J1856.2+0450c is fit to an extension of $\sim2\degree$
and the result is statistically significant with \tsext=45.4. However,
by looking at the residuals it is clear that this complicated region is
not well modeled. We manually discard sources like this.
% information came from
% /nfs/slac/g/ki/ki03/lande/extended_catalog/2FGL/v15/standard_analysis/spectral_emin_1000_v1/P72Y3047/v1/results_followup_P72Y3047.yaml
% (2FGL J1856.2+0450c = P72Y3047 for)
